Reliable error estimation for Sobol' indices

### 2016-08-12 article.tex
changes L

% !TEX TS-program = PDFLatexBibtex
%&LaTeX
\documentclass[]{elsarticle}
\setlength{\marginparwidth}{0.5in}
\usepackage{amsmath,amssymb,amsthm,natbib,
mathtools,bbm,extraipa,mathabx,graphicx,algorithm}
\usepackage{algpseudocode}
%accents,

@@

\subsection{Definition of $\widehat{S}$ (fix it with max and min)}
In some applications, it is more useful to estimate the normalized Sobol' indices, $S_u = \underline{\tau}_u^2/\sigma^2$ and $\overline{S}_u = \overline{\tau}_u^2/\sigma^2$. In both cases, regular and normalized indices can be approximated by estimating some integrals,
\begin{align*}
\underline{S}_u & = \frac{\int_{[0,1)^{2d-1}} f(\vx)f(\vx_{u}:{\vx'}_{-u})d\vx d{\vx'} - \mu^2 }{\int_{[0,1)^{d}} f(\vx)^2 d{\vx}-\mu^2} = \frac{I_1-(I_3)^2}{I_2-(I_3)^2}, \\
\overline{S}_u & = \frac{\frac{1}{2}\int_{[0,1)^{d+1}}(f(\vx')-f(\vx_u:{\vx'}_{-u}))^2d\vx d{\vx'}}{\int_{[0,1)^{d}} f(\vx)^2 d{\vx}-\mu^2} = \frac{I_1}{I_2-(I_3)^2}.
\end{align*}
Therefore, we can define $\underline{S}_u(\vI)$ and $\overline{S}_u(\vI)$ as functions over a vector of integrals $\vI$. If we can estimate $\vI$ such that $\vI\in B_{\varepsilon_{\vI}}(\widehat{\vI}):=[\widehat{\vI}-\varepsilon_{\vI},\widehat{\vI}+\varepsilon_{\vI} ]$, then we can define the following estimators for the Sobol' indices:
\begin{align*}
\widehat{S}_u & = \frac{1}{2}\left(\min\left(\max_{\vI\in B_{\varepsilon_{\vI}}(\widehat{\vI})} S_u(\vI),1\right) + \max\left(\min_{\vI\in B_{\varepsilon_{\vI}}(\widehat{\vI})} S_u(\vI),0\right) \right) \\
\varepsilon_{S_u} & = \frac{1}{2}\left(\min\left(\max_{\vI\in B_{\varepsilon_{\vI}}(\widehat{\vI})} S_u(\vI),1\right) - \max\left(\min_{\vI\in B_{\varepsilon_{\vI}}(\widehat{\vI})} S_u(\vI),0\right) \right)
\end{align*}
where the true value $S_u\in [\widehat{S}_u - \varepsilon_{S_u}, \widehat{S}_u + \varepsilon_{S_u}]$.

@@

Algorithm \ref{recursive.algorithm} summarizes the main steps of our recursive procedure. First, fix the target precision $\epsilon >0$. Set $m=m_0$ and construct the two designs $\mathcal{P}_{m}=\{\vx_i\}_{i=0}^{2^{m}-1}$ and $\mathcal{P}'_{m}=\{\vx'_i\}_{i=0}^{2^{m}-1}$ accordingly to the multiplicative approach detailed in \cite{crass}. With this construction, $\mathcal{P}_{\ell}$ and $\mathcal{P}'_{m}$ correspond to the first $2^{m}$ points of two Sobol' sequences. Furthermore, they possess the structure of two replicated designs of order $1$.
\bigskip

Then, $\mathcal{P}_{m}$ and $\mathcal{P}'_{m}$ can either be used with Saltelli's strategy to estimate all first-order indices and total effect Sobol' indices. This option is referred as Variant $A$ in Algorithm \ref{recursive.algorithm}. Or $\mathcal{P}_{m}$ and $\mathcal{P}'_{m}$ can be used with the replication procedure to estimate all first-order Sobol' indices. This option is referred as Variant $B$ in Algorithm \ref{recursive.algorithm}.

@@

The cost of our algorithm varies whether Variant $A$ or Variant $B$ is selected. To discuss this cost we note by $m^\star$ the ending iteration. If Variant $A$ is selected, the cost of our algorithm writes:
\begin{equation*}
\sum \limits_{u \in \mathcal{D}} 2^{m_u} + 2 \times 2^{m^{\star}}, \qquad m^\star= \max \limits_{u \in \mathcal{D}} \ m_u,
\end{equation*}
where: 
\begin{itemize}
\item[$\bullet$] $2^{m_u}$ is the number of evaluations $f\left(\vx_{i,u}:{\vx'}_{i,-u}\right)$ used to estimate both $\underline{S}_u$ and $\overline{S}_u$.
\item[$\bullet$] $2 \times 2^{m^{\star}}$ is the number of evaluations $f\left(\vx_{i}\right)$ and $f\left({\vx'}_{i}\right)$ used in the estimation of each first-order and total effect index.
\end{itemize}
This cost can be seen as a generalisation of the one specified in Theorem \ref{saltelli.theorem}. If all $m_u$ are equal, the cost of Variant $A$ becomes $2^{m^\star}(d+2)$ and we recover the cost of Theorem \ref{saltelli.theorem} with $n=2^{m^\star}$.

@@

The cost of Variant $A$ including ``Correlation 2" writes: 
\begin{equation}
\sum \limits_{j \in \Gamma} 2^{m''_{u_j}} + \sum \limits_{j \in \Gamma} 2^{m'_{u_j}} + \sum \limits_{j \in \mathcal{D}/\Gamma} 2^{m_{u_j}} + 2 \times 2^{m^\star}, \qquad m'_{u_j}\leq m^{\star}, \ m_{u_j} \leq m^{\star},
\label{cost.improvement}
\end{equation}
where: \begin{itemize}
\item[$\bullet$] for $j\in \Gamma$, $2^{m''_{u_j}}$ is the number of evaluations $f\left(\vz_{i,u_j}:\vx_{i,-u_j}\right)$ to estimate small first-order indices,
\item[$\bullet$] for $j\in \Gamma$, $2^{m'_{u_j}}$ is the number of evaluations $f\left(\vx_{i,j}:{\vx'}_{i,-j}\right)$ to estimate both small first-order and total effect indices,
\item[$\bullet$] for $j\in \mathcal{D}/\Gamma$, $ 2^{m_{u_j}}$ is the number of evaluations $f\left(\vx_{i,j}:{\vx'}_{i,-j}\right)$ to estimate non small first-order indices,
\item[$\bullet$] $2 \times 2^{m^{\star}}$ is the number of evaluations $f\left(\vx_{i}\right)$ and $f\left({\vx'}_{i}\right)$ used in the estimation of each first-order and total effect index.
\end{itemize}
Recall that the cost of Variant $A$ without  ``Correlation 2" writes:
\begin{equation}
\sum \limits_{j \in \Gamma} 2^{m_{u_j}} + \sum \limits_{j \in \mathcal{D}/\Gamma} 2^{m_{u_j}} + 2 \times 2^{m^{\star}}, \qquad m_{u_j} \leq m^{\star}.
\label{cost.non.improvement}
\end{equation}
The difference between costs (\ref{cost.improvement}) and (\ref{cost.non.improvement}) equals:
\begin{equation}
 \sum \limits_{j \in \Gamma} 2^{m_{u_j}} \left(2^{m''_{u_j}-m_{u_j}}  + 2^{m'_{u_j}-m_{u_j}} - 1 \right) = \sum \limits_{j \in \Gamma} c_j.
\label{cost.comparison}
\end{equation}
Hence, the sign of this difference indicates whether or not using ``Correlation 2" bring an improvement to Variant $A$. We distinguish two cases :
\begin{itemize}
\item[1)] for $j \in \Gamma$, the total effect index $\ov{$S$}_j$ requires as much or more evaluations $f\left(\vx_{i,u_j}:{\vx'}_{i,-u_j}\right)$ than the first-order index $\underline{S}_j$. Since the total effect estimator is the same, as a consequence we have $m'_{u_j}=m_{u_j}$ and $c_j >0$.
\item[2)] for $j \in \Gamma$, the total effect index $\ov{$S$}_j$ requires less evaluations $f\left(\vx_{i,u_j}:{\vx'}_{i,-u_j}\right)$ than the first-order index $\underline{S}_j$. In this case, if both $m''_{u_j} < m_{u_j}$ and $m'_{u_j} < m_{u_j}$ then $c_j \leq 0$.  
\end{itemize}
We can write $\Gamma = \Gamma_1 \cup \Gamma_2$, $\Gamma_1 \cap \Gamma_2=\varnothing$, to separate the number of times case $1)$ or case $2)$ are verified. We expect to have more case $2)$ than case $1)$. The reason is that the numerator of the first-order index (\ref{}) requires to estimate $2d-1$ integrals against only $d+1$ integrals for the numerator of the total effect index (\ref{}). Thus, it seems reasonable to suppose that it takes less points to evaluate (\ref{}) than (\ref{}). 
Furthermore, in case $2$, we expect the two conditions $m''_{u_j} < m_{u_j}$ and $m'_{u_j} < m_{u_j}$ to be often satisfied as ``Correlation 2" has been shown 
to perform the best.

To support this discussion, illustrations of the use of ``Correlation 2" are presented in Section \ref{appli}. 
\bigskip

In pratice, one does not know which are the small Sobol' indices. To overcome this issue, we propose the following alternative for Variant $A$. Assume that there exists one or multiple  $u \in \mathcal{D}$ such that estimates $\widehat{\underline{S}_u}^{(m)}$ are smaller than a specified threshold at the end of the first iteration of Algorithm \ref{recursive.algorithm}. Then, for these indices, at the next iteration estimator (\ref{first.order.est}) is switched for (\ref{correlation2}) and a third Sobol' sequence $\mathcal{P''}_{m}=\{\vz_i\}_{i=0}^{2^m-1}$ is constructed for the evaluations $f({\vz}_{i,u}:{\vx}_{i,-u})$.

>>>>

\documentclass[]{elsarticle}
\setlength{\marginparwidth}{0.5in}
\usepackage{amsmath,amssymb,amsthm,natbib,
mathtools,bbm,extraipa,mathabx,graphicx,algorithm}
\usepackage{algpseudocode}
%accents,

@@

\subsection{Definition of $\widehat{S}$ (fix it with max and min)}
In some applications, it is more useful to estimate the normalized Sobol' indices, $S_u = \underline{\tau}_u^2/\sigma^2$ and $\overline{S}_u = \overline{\tau}_u^2/\sigma^2$. In both cases, regular and normalized indices can be approximated by estimating some integrals,
\begin{align*}
\underline{S}_u & = \frac{\int_{[0,1)^{2d-1}} f(\vx)f(\vx_{u}:{\vx'}_{-u})d\vx d{\vx'} - \mu^2 }{\int_{[0,1)^{d}} f(\vx)^2 d{\vx}-\mu^2} = \frac{I_1-(I_3)^2}{I_2-(I_3)^2}, \\
\overline{S}_u & = \frac{\frac{1}{2}\int_{[0,1)^{d+1}}(f(\vx')-f(\vx_u:{\vx'}_{-u}))^2d\vx d{\vx'}}{\int_{[0,1)^{d}} f(\vx)^2 d{\vx}-\mu^2} = \frac{I_1}{I_2-(I_3)^2}.
\end{align*}
Therefore, we can define $\underline{S}_u(\vI)$ and $\overline{S}_u(\vI)$ as functions over a vector of integrals $\vI$. If we can estimate $\vI$ such that $\vI\in B_{\varepsilon_{\vI}}(\widehat{\vI}):=[\widehat{\vI}-\varepsilon_{\vI},\widehat{\vI}+\varepsilon_{\vI} ]$, then we can define the following estimators for the Sobol' indices:
\begin{align*}
\widehat{S}_u & = \frac{1}{2}\left(\min\left(\max_{\vI\in B_{\varepsilon_{\vI}}(\widehat{\vI})} S_u(\vI),1\right) + \max\left(\min_{\vI\in B_{\varepsilon_{\vI}}(\widehat{\vI})} S_u(\vI),0\right) \right) \\
\varepsilon_{S_u} & = \frac{1}{2}\left(\min\left(\max_{\vI\in B_{\varepsilon_{\vI}}(\widehat{\vI})} S_u(\vI),1\right) - \max\left(\min_{\vI\in B_{\varepsilon_{\vI}}(\widehat{\vI})} S_u(\vI),0\right) \right)
\end{align*}
where the true value $S_u\in [\widehat{S}_u - \varepsilon_{S_u}, \widehat{S}_u + \varepsilon_{S_u}]$.
\begin{figure}[!ht]
\centering
\includegraphics{scheme_S.eps}
\end{figure}

@@

Algorithm \ref{recursive.algorithm} summarizes the main steps of our recursive procedure. First, fix the target precision $\epsilon >0$. Set $m=m_0$ and construct the two designs $\mathcal{P}_{m}=\{\vx_i\}_{i=0}^{2^{m}-1}$ and $\mathcal{P}'_{m}=\{\vx'_i\}_{i=0}^{2^{m}-1}$ accordingly to the multiplicative approach detailed in \cite{crass}. With this construction, $\mathcal{P}_{m}$ and $\mathcal{P}'_{m}$ correspond to the first $2^{m}$ points of two Sobol' sequences. Furthermore, they possess the structure of two replicated designs of order $1$.
\bigskip

$\mathcal{P}_{m}$ and $\mathcal{P}'_{m}$ can either be used with Saltelli's strategy to estimate all first-order indices and total effect Sobol' indices. This option is referred as Variant $A$ in Algorithm \ref{recursive.algorithm}. Or $\mathcal{P}_{m}$ and $\mathcal{P}'_{m}$ can be used with the replication procedure to estimate all first-order Sobol' indices. This option is referred as Variant $B$ in Algorithm \ref{recursive.algorithm}.

@@

The cost of our algorithm varies whether Variant $A$ or Variant $B$ is selected. To discuss this cost we note by $m^\star$ the ending iteration. If Variant $A$ is selected, the cost of our algorithm writes:
\begin{equation*}
\sum \limits_{u \in \mathcal{D}} 2^{m_u} + 2 \times 2^{m^{\star}}, \qquad m^\star= \max \limits_{u \in \mathcal{D}} \ m_u,
\end{equation*}
where: 
\begin{itemize}
\item[$\bullet$] $2^{m_u}$ is the number of evaluations $f\left(\vx_{i,u}:{\vx'}_{i,-u}\right)$ used to estimate both the first-order index $\underline{S}_u$ and the total effect index $\overline{S}_u$.
\item[$\bullet$] $2 \times 2^{m^{\star}}$ is the number of evaluations $f\left(\vx_{i}\right)$ and $f\left({\vx'}_{i}\right)$ used in the estimation of each first-order and total effect index.
\end{itemize}
This cost can be seen as a generalisation of the one specified in Theorem \ref{saltelli.theorem}. If all $m_u$ are equal, the cost of Variant $A$ becomes $2^{m^\star}(d+2)$ and we recover the cost in Theorem \ref{saltelli.theorem} with $n=2^{m^\star}$.

@@

The cost of Variant $A$ including ``Correlation 2" writes: 
\begin{equation}
\sum \limits_{j \in \Gamma} 2^{m''_{u_j}} + \sum \limits_{j \in \Gamma} 2^{m'_{u_j}} + \sum \limits_{j \in \mathcal{D}/\Gamma} 2^{m_{u_j}} + 2 \times 2^{m^\star}, \qquad m'_{u_j}\leq m^{\star}, \ m_{u_j} \leq m^{\star},
\label{cost.improvement}
\end{equation}
where: \begin{itemize}
\item[$\bullet$] for $j\in \Gamma$, $2^{m''_{u_j}}$ is the number of evaluations $f\left(\vz_{i,u_j}:\vx_{i,-u_j}\right)$ to estimate $\underline{S}_{u_j}$,
\item[$\bullet$] for $j\in \Gamma$, $2^{m'_{u_j}}$ is the number of evaluations $f\left(\vx_{i,j}:{\vx'}_{i,-j}\right)$ to estimate both $\underline{S}_{u_j}$ and  $\overline{S}_{u_j}$,
\item[$\bullet$] likewise, for $j\in \mathcal{D}/\Gamma$, $ 2^{m_{u_j}}$ is the number of evaluations $f\left(\vx_{i,j}:{\vx'}_{i,-j}\right)$ to estimate both $\underline{S}_{u_j}$ and  $\overline{S}_{u_j}$,
\item[$\bullet$] $2 \times 2^{m^{\star}}$ is the number of evaluations $f\left(\vx_{i}\right)$ and $f\left({\vx'}_{i}\right)$ used in the estimation of each first-order and total effect index.
\end{itemize}
Recall that the cost of Variant $A$ without  ``Correlation 2" writes:
\begin{equation}
\sum \limits_{j \in \Gamma} 2^{m_{u_j}} + \sum \limits_{j \in \mathcal{D}/\Gamma} 2^{m_{u_j}} + 2 \times 2^{m^{\star}}, \qquad m_{u_j} \leq m^{\star}.
\label{cost.non.improvement}
\end{equation}
The difference between costs (\ref{cost.improvement}) and (\ref{cost.non.improvement}) equals:
\begin{equation}
 \sum \limits_{j \in \Gamma} 2^{m_{u_j}} \left(2^{m''_{u_j}-m_{u_j}}  + 2^{m'_{u_j}-m_{u_j}} - 1 \right) = \sum \limits_{j \in \Gamma} c_j.
\label{cost.comparison}
\end{equation}
Hence, the sign of this difference indicates whether or not using ``Correlation 2" bring an improvement to Variant $A$. We distinguish two cases :
\begin{itemize}
\item[1)] for $j \in \Gamma$, the total effect index $\ov{$S$}_{u_j}$ requires as much or more evaluations $f\left(\vx_{i,u_j}:{\vx'}_{i,-u_j}\right)$ than the first-order index $\underline{S}_{u_j}$. Since the total effect estimator is the same, as a consequence we have $m'_{u_j}=m_{u_j}$ and $c_j >0$.
\item[2)] for $j \in \Gamma$, the total effect index $\ov{$S$}_{u_j}$ requires less evaluations $f\left(\vx_{i,u_j}:{\vx'}_{i,-u_j}\right)$ than the first-order index $\underline{S}_{u_j}$. In this case, if both $m''_{u_j} < m_{u_j}$ and $m'_{u_j} < m_{u_j}$ then $c_j \leq 0$.  
\end{itemize}
We can write $\Gamma = \Gamma_1 \cup \Gamma_2$, $\Gamma_1 \cap \Gamma_2=\varnothing$, to separate the number of times case $1)$ or case $2)$ are verified. We expect to have more case $2)$ than case $1)$. Indeed, the numerator of the first-order index $\underline{S}_u$ requires to estimate $2d-1$ integrals against only $d+1$ integrals for the numerator of the total effect index $\ov{$S$}_u$. It seems reasonable to suppose that it takes less points to evaluate $\ov{$S$}_u$ than $\underline{S}_u$.

Furthermore, in case $2)$, we expect the two conditions $m''_{u_j} < m_{u_j}$ and $m'_{u_j} < m_{u_j}$ to be often satisfied as ``Correlation 2" has been shown 
to perform the best for small first-order indices. To support this discussion, illustrations of the use of ``Correlation 2" are presented in Section \ref{appli}. 
\bigskip

In pratice, one does not know which are the small Sobol' indices. To overcome this issue, we propose the following alternative for Variant $A$. If at the end of the first iteration, the estimate $\widehat{\underline{S}_u}^{(m)}$, $u \in \mathcal{D}$, is smaller than a specified threshold, then at the next iteration estimator (\ref{first.order.est}) is switched for (\ref{correlation2}) and a third Sobol' sequence $\mathcal{P''}_{m}=\{\vz_i\}_{i=0}^{2^m-1}$ is constructed for the evaluations $f({\vz}_{i,u}:{\vx}_{i,-u})$.
